---
title: "Colombia: Event Data Validation Project, Data Codebook"
thanks: "`r sprintf('Document date: %s', format(Sys.time(), '%B %d, %Y'))`"
author:
- name: Logan Stundal, Ban Bagozzi, John Freeman, Jen Holmes
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: false
    fig_caption: true
    latex_engine: pdflatex
    template: "c:/users/logan/googledrive/umn/research/templates_logos/templates/svm-r-markdown-templates-master/svm-latex-ms_logan.tex"   
bibliography: "c:/users/logan/googledrive/umn/research/references/library"
biblio-style: "c:/users/logan/googledrive/umn/research/references/csl-formats/apsrnourl.bst"
header-includes:
  -  \usepackage{hyperref}
  -  \usepackage{amssymb}
  -  \usepackage{caption}
  -  \usepackage{subfig}
  -  \usepackage{multirow}
  -  \usepackage{booktabs}
  -  \usepackage{tabularx}
geometry: margin=1in
fontfamily: mathpazo
fontsize: 12pt
spacing: onehalf
endnote: no
logan: true
linestretch: 1.5
citecolor: black
linkcolor: black
toc: false
---
\vspace{3mm}

```{r include = FALSE}
library(knitr)
library(kableExtra)
library(sp)
library(ggplot2)

opts_chunk$set(echo = FALSE)

load('c:/users/logan/googledrive/umn/research/ra_john/event_data_project/data/colombia.rdata')
```

# Table of variables
- Include a nicely formatted table of *variables*, *variable names in data*, *descriptions*, and *units* here.

\begin{table}[!h]
  \begin{tabularx}{\textwidth}{lXXXr}
    \toprule
    Variable & Name & Description & Unit & Source \\
    \midrule
    Var 1 & Var 1 & Var 1 & Var 1 & Var 1 \\
    \bottomrule
  \end{tabularx}
\end{table}


OR -----

```{r}

d <- data.frame('Variable'  = c('a','b','c'),
                'Data Name' = c('bob','bill','other bill'),
                'Description' = c(letters[1:3]),
                'Units'       = c(letters[1:3]))

kable(d) %>%
  kable_styling(bootstrap_options = "striped", full_width = T)

```



The following sections provide more detailed information on these variable constructions.

\newpage
# Units of Analysis and Shapefile
The shapefile for our Colombia data come from the United Nations Office for the Coordination of Humanitarian Affairs, hosted on the Humanitarian Data Exchange: [HDX](https://data.humdata.org/dataset/colombia-administrative-boundaries-levels-0-3). The file contains coordinate information for Colombia's 1117 municipalities. San Andres and Providencia, a department in the Carribean, was dropped from this base map file due to their distance from mainland Colombia resulting in a final base map of 1116 municipalities[^gadm]. 

```{r, cache=TRUE, fig.cap='Colombia, Level 2 Municipalities (n = 1116)',fig.height=4,fig.width=4}
ggplot(data = colombia) + geom_sf(fill = 'transparent', size = 0.1) + theme_void()

```


[^gadm]: GADM, a repository hosting spatial data on administrative units, has [errors](https://gis.stackexchange.com/questions/167247/seeking-colombia-municipality-geometries) in its version of Colombia's municipalities map.

## General conditions:
This shapefile was reprojected into a XXXX projection in order to accurately estimate area size and all distance measures.

All variables in the data were aggregated to cover years betten **2002** and **2009** .

\newpage
# Event Data
## ICEWS

## GED

## CINAP


# Covariate Data
## PRIO
### Mountains Mean
### Forest Mean


## Google Earth Engine
### World Population
### Forest Cover
### Night Lights

### Rough Terrain Index
Note, this sources is from xyz.


## Electoral Data
- source: Constituency-Level Elections Archive  [CLEA] Lower Chamber Elections Archive [Version: June 17, 2019]
  - Link: http://www.electiondataarchive.org/clea-lower-chamber-elections-archive.php
